---
title: "FINC-580 Assignment #2 — Dual Moving Average Crossover Strategy"
author: "Huy Le (hl9082)"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This script implements a Dual Moving Average Crossover strategy using Tesla (TSLA) stock.

The code calculates trading signals, portfolio performance, Sharpe ratio, and maximum drawdown for the 2019-2024 period.

An extra credit section replicates the analysis using a machine learning K-Nearest Neighbors (KNN) model.

### 0. Setting up libraries and preparing helper function
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# 1. Define ALL the packages you need (now including 'caret' for the Extra Credit)
required_packages <- c("quantmod", "PerformanceAnalytics", "ggplot2", "tidyr", "caret", "dplyr")
install.packages("caret", repos = "http://cran.rstudio.com/", dependencies = TRUE)

# 2. Check which ones are NOT already installed on your computer
new_packages <- required_packages[!(required_packages %in% installed.packages()[, "Package"])]

# 3. If any are missing, install them automatically
if (length(new_packages) > 0) {
  install.packages(new_packages,
                   dependencies = TRUE,
                   repos = "[https://cloud.r-project.org](https://cloud.r-project.org)")
}
# Load required packages
library(quantmod)
library(PerformanceAnalytics)
library(ggplot2)
library(tidyr)
library(caret)
library(dplyr)

# ---- Helper Function to Display First and Last Rows ----
head.tail <- function(x) rbind(head(x), tail(x))
```


### 1. Technical Implementation & Visualization
Data Collection & Moving Averages
Stock Selected: Tesla

Data Period: January 1, 2019, to December 31, 2024.

We utilize quantmod to download the Adjusted Closing prices and calculate the 20-day (Fast) and 100-day (Slow) Simple Moving Averages.

```{r}
ticker <- "TSLA" 
getSymbols(ticker, from = "2019-01-01", to = "2024-12-31", periodicity = "daily")

# Extract Adjusted Close
price <- Ad(get(ticker))
colnames(price) <- "Price"

head.tail(price)
```
## Calculate 20-Day and 100-Day Moving Averages
```{r}
# Calculate SMAs using TTR package (loaded with quantmod)
sma20 <- SMA(price, n = 20)
sma100 <- SMA(price, n = 100)

# Combine into a single xts object
data <- merge(price, sma20, sma100)
colnames(data) <- c("Price", "SMA20", "SMA100")
data <- na.omit(data) # Remove the first 100 days where SMA100 cannot be calculated
head.tail(data)
```

## Plot Price and Moving Averages:
```{r}
# Convert to dataframe for ggplot
plot_data <- data.frame(Date = index(data), coredata(data))
plot_data_long <- pivot_longer(plot_data, cols = -Date, names_to = "Series", values_to = "Value")

ggplot(plot_data_long, aes(x = Date, y = Value, color = Series)) +
  geom_line(linewidth = 0.8) +
  scale_color_manual(values = c("Price" = "black", "SMA20" = "blue", "SMA100" = "red")) +
  labs(title = paste(ticker, "- Price vs. 20-Day and 100-Day SMA"),
       y = "Price ($)", x = "Date") +
  theme_minimal()
```

### 2. Trading Signal Generation and Portfolio Return:
Generating the Trading Signal
- Rule: If SMA20 > SMA100, Invested = 1. Else, Invested = 0.
- Lag: We lag the signal by exactly 1 period lag() to prevent look-ahead bias (we trade the day after the signal is generated).

```{r}
# Create the signal: 1 if SMA20 > SMA100, 0 otherwise
data$Signal <- ifelse(data$SMA20 > data$SMA100, 1, 0)

# Lag the signal by 1 day!
data$Position <- quantmod::Lag(data$Signal, k = 1)
data <- na.omit(data)

head.tail(data[, c("SMA20", "SMA100", "Signal", "Position")])
```
## Calculate Portfolio Returns
```{r}
# Calculate daily returns of the stock
data$Stock_Return <- ROC(data$Price, type = "discrete")

# Calculate Strategy Returns (Stock Return * Position)
data$Strategy_Return <- data$Stock_Return * data$Position
data <- na.omit(data)

# Calculate cumulative growth of $1 for both Strategy and Buy & Hold
data$Cum_Strategy <- cumprod(1 + data$Strategy_Return)
data$Cum_BnH <- cumprod(1 + data$Stock_Return)

head.tail(data[, c("Stock_Return", "Strategy_Return", "Cum_Strategy", "Cum_BnH")])
```
## Plot Cumulative Portfolio Value
```{r}
# Plotting the comparison
plot_cum <- data.frame(Date = index(data), 
                       Strategy = coredata(data$Cum_Strategy), 
                       Buy_And_Hold = coredata(data$Cum_BnH))
plot_cum_long <- pivot_longer(plot_cum, cols = -Date, names_to = "Portfolio", values_to = "Value")

# Calculate the maximum y-value so we know how high to draw the chart
max_y <- ceiling(max(plot_cum_long$Value, na.rm = TRUE))
ggplot(plot_cum_long, aes(x = Date, y = Value, color = Portfolio)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", color = "black", linewidth = 0.8) +
   
  scale_y_continuous(limits = c(0, max_y), breaks = unique(c(0, 1, seq(0, max_y, by = 5)))) +
  scale_color_manual(values = c("Cum_BnH" = "black", "Cum_Strategy" = "purple"),
                     labels = c("Cumulative Buy & Hold", "Cumulative Strategy")) +
  labs(title = "Cumulative Returns: Dual SMA Strategy vs. Buy & Hold",
       y = "Value of $1 Investment", x = "Date") +
  theme_minimal()
```

### 3. Performance Evaluation
We evaluate the risk-adjusted performance using the Sharpe Ratio (Risk-Free Rate = 0.02) and Maximum Drawdown.
```{r}
rf_daily <- 0.02 / 252 # Convert annual RF rate to daily

# Annualized Sharpe Ratio
sharpe_strat <- SharpeRatio.annualized(data$Strategy_Return, Rf = rf_daily)
sharpe_bnh <- SharpeRatio.annualized(data$Stock_Return, Rf = rf_daily)

# Maximum Drawdown
maxdd_strat <- maxDrawdown(data$Strategy_Return)
maxdd_bnh <- maxDrawdown(data$Stock_Return)

# Print cleanly
cat("Strategy Annualized Sharpe Ratio:", round(sharpe_strat, 4), "\n")
cat("Buy & Hold Annualized Sharpe Ratio:", round(sharpe_bnh, 4), "\n\n")

cat("Strategy Max Drawdown:", round(maxdd_strat * 100, 2), "%\n")
cat("Buy & Hold Max Drawdown:", round(maxdd_bnh * 100, 2), "%\n")
```

### 4. Analysis
Q: Compare the final cumulative value and comment on whether the strategy was successful based on the Sharpe Ratio.

Based on the cumulative returns chart, the final value of a $1 investment in the SMA20/SMA100 strategy was $X.XX, compared to $Y.YY for the standard buy-and-hold approach. (Did it beat the market?)

Looking at risk-adjusted performance, the Strategy generated a Sharpe Ratio of 0.7457, compared to the Buy & Hold Sharpe of 1.1172. Furthermore, the strategy limited our Maximum Drawdown to 35.38%, while holding the stock exposed us to a worse drawdown of 31.43%. Ultimately, the strategy was [successful / unsuccessful] because... (explain if the reduced risk/drawdown was worth any potential underperformance in total return!)

---

### 5. Extra Credit: K-Nearest Neighbors (KNN) Trading Strategy

## Theoretical Framework & Feature Engineering
As discussed in the Ang textbook (Chapter 12), while simple moving averages are rigid heuristic rules, machine learning algorithms like K-Nearest Neighbors (KNN) allow the data to dictate the optimal pattern recognition. 

KNN is a non-parametric classification algorithm. We will use it to predict the **direction of tomorrow's return** (Up or Down) based on recent historical price momentum (our features). 

To prevent the model from "peeking into the future," our predictive features for Day $t$ MUST be strictly lagged (e.g., returns from Day $t-1$, $t-2$, and $t-5$). We cannot use today's closing price to predict today's return!

## KNN Data Preparation & Feature Engineering
```{r knn-setup}
# 1. Start fresh with the 'price' object from Part 1
# Calculate daily returns natively as an xts time-series
knn_rets <- ROC(price, type = "discrete")

# 2. Create lagged features using quantmod's Lag function directly on the xts object.
# Doing this in 'xts' format first avoids ALL dataframe column/NULL errors and 
# perfectly aligns the dates mathematically.
knn_xts <- merge(
  Return = knn_rets,
  Lag1 = quantmod::Lag(knn_rets, k = 1),
  Lag2 = quantmod::Lag(knn_rets, k = 2),
  Lag5 = quantmod::Lag(knn_rets, k = 5)
)

# Force exact column names so we know exactly what they are
colnames(knn_xts) <- c("Return", "Lag1", "Lag2", "Lag5")

# 3. NOW convert the perfectly aligned time-series into a standard dataframe 
# for the Machine Learning algorithm to read.
knn_df <- data.frame(Date = index(knn_xts), coredata(knn_xts))

# 4. Create the target classification variable
knn_df$Target <- as.factor(ifelse(knn_df$Return > 0, "Up", "Down"))

# 5. Remove all rows with NAs (due to the 5-day lag and initial ROC calculation)
knn_df <- na.omit(knn_df)

head.tail(knn_df)
```

## Train the KNN Model
- Model Training with K-Fold Cross-ValidationTo prevent overfitting (creating a model that perfectly memorizes the past but fails in the future), we will use k-fold cross-validation. 
- The caret package will split our data into 5 separate chunks (folds). It will train the model on 4 folds, test it on the 1 remaining fold, and repeat this process to find the mathematically optimal number of "neighbors" (the $k$ in KNN) to base its predictions on.
```{r}
# 1. Set a random seed so the cross-validation split is strictly reproducible
set.seed(580)
# 2. Define the training control parameters: 5-fold cross-validation
train_control <- trainControl(method = "cv", number = 5)
# 3. Train the K-Nearest Neighbors model
# Formula: Target ~ Lag1 + Lag2 + Lag5 (Predict Target using our 3 lagged features)
knn_model <- train(Target ~ Lag1 + Lag2 + Lag5, 
                   data = knn_df, 
                   method = "knn",
                   trControl = train_control,
                   preProcess = c("center", "scale"), # Standardizes data (crucial for distance-based KNN!)
                   tuneLength = 5) # Tells caret to test 5 different values of 'k'
# Print the model summary to see which 'k' (number of neighbors) performed best
print(knn_model)
```

## Backtest the KNN Strategy
```{r}
# Generate predictions
knn_predictions <- predict(knn_model, newdata = knn_df)

# Translate predictions into trading signals (1 = Invested, 0 = Cash)
knn_df$Position <- ifelse(knn_predictions == "Up", 1, 0)

# Calculate Daily Returns & Cumulative Returns
knn_df$Strategy_Return <- knn_df$Return * knn_df$Position
knn_df$Cum_Strategy <- cumprod(1 + knn_df$Strategy_Return)
knn_df$Cum_BnH <- cumprod(1 + knn_df$Return)

head.tail(knn_df[, c("Target", "Position", "Cum_Strategy", "Cum_BnH")])
```

## Evaluate KNN Performance
- We will now evaluate if the machine learning model managed to outperform both the standard Buy-and-Hold approach and the rigid Dual SMA crossover strategy we built in Part 2.
```{r}
# --- PLOT ---
plot_knn <- pivot_longer(knn_df[, c("Date", "Cum_Strategy", "Cum_BnH")], 
                         cols = -Date, names_to = "Portfolio", values_to = "Value")

plot(ggplot(plot_knn, aes(x = Date, y = Value, color = Portfolio)) +
  geom_line(linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_color_manual(values = c("Cum_BnH" = "black", "Cum_Strategy" = "purple"),
                     labels = c("Buy & Hold", "KNN ML Strategy")) +
  labs(title = "Machine Learning (KNN) Strategy vs. Buy & Hold",
       y = "Value of $1 Investment", x = "Date") +
  theme_minimal())

# --- METRICS ---
# Strip all complex formatting to ensure clean numeric and Date data
safe_dates <- as.Date(as.character(knn_df$Date))
safe_strat_ret <- as.numeric(knn_df$Strategy_Return)
safe_bnh_ret <- as.numeric(knn_df$Return)

# Create the xts objects safely
strat_returns_xts <- xts(safe_strat_ret, order.by = safe_dates)
bnh_returns_xts <- xts(safe_bnh_ret, order.by = safe_dates)

# Calculate Risk-Free Rate dynamically
rf_daily <- 0.02 / 252 

knn_sharpe <- SharpeRatio.annualized(strat_returns_xts, Rf = rf_daily)
bnh_sharpe <- SharpeRatio.annualized(bnh_returns_xts, Rf = rf_daily)
knn_maxdd <- maxDrawdown(strat_returns_xts)
bnh_maxdd <- maxDrawdown(bnh_returns_xts)

cat("--- KNN STRATEGY vs BUY & HOLD ---\n")
cat(sprintf("KNN Strategy Sharpe Ratio: %.4f\n", knn_sharpe))
cat(sprintf("Buy & Hold Sharpe Ratio:   %.4f\n\n", bnh_sharpe))
cat(sprintf("KNN Strategy Max Drawdown: %.2f%%\n", knn_maxdd * 100))
cat(sprintf("Buy & Hold Max Drawdown:   %.2f%%\n", bnh_maxdd * 100))
```

## KNN Strategy Analysis
[Student Answer Placeholder] >
Write a brief paragraph here based on the numbers that print out! Did the machine learning model beat the moving average strategy from Part 2? Why do you think it did or didn't? Focus heavily on how the KNN model attempts to classify volatility based on lagged historical patterns, whereas the SMA is a lagging trend-follower. Note if the Maximum Drawdown was significantly reduced by the model sitting in cash during periods classified as "Down" trends.