---
title: "FINC-580 Assignment #1 — Portfolio Returns and Optimization"
author: "Huy"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
# Load necessary libraries
# Check for missing packages and install them if necessary
# 1. Define the packages you need
required_packages <- c("quantmod", "PerformanceAnalytics", "ggplot2", "tidyr")

# 2. Check which ones are NOT already installed on your computer
new_packages <- required_packages[
                                  !(required_packages %in%
                                      installed.packages()[, "Package"])]

# 3. If any are missing, install them using the standard global mirror
if (length(new_packages) > 0) {
  install.packages(new_packages,
                   dependencies = TRUE,
                   repos = "https://cloud.r-project.org")
}

# 4. Load the packages
library(quantmod)
library(PerformanceAnalytics)
library(ggplot2)
library(tidyr)
```
### Assignment Overview
This assignment introduces portfolio construction, 
return calculation, and optimization. 
We collect real-world stock price data to 
construct an equal-weighted portfolio 
and analyze its performance 
under a buy-and-hold strategy. 
We also explore modern portfolio theory by 
plotting the efficient frontier, 
identifying the minimum-variance portfolio, 
and determining the optimal risky portfolio.

# Part 1: Portfolio Return Calculation
1. Stock Selection & Data Collection
We selected five publicly traded stocks across different sectors. 
We use the quantmod package to download adjusted 
closing prices from January 2021 to September 2025.
```{r}
# Define 5 stock tickers
tickers <- c("AAPL", "MSFT", "JNJ", "XOM", "JPM")
start_date <- "2021-01-01"
end_date <- "2025-09-30"

# Download adjusted closing prices
getSymbols(tickers,
           src = "yahoo",
           from = start_date,
           to = end_date,
           periodicity = "monthly")

# Combine adjusted prices into one xts object and rename columns
prices <- do.call(merge, lapply(tickers, function(x) Ad(get(x))))
colnames(prices) <- tickers
prices <- na.omit(prices)
```
2. Construct an Equal-Weighted Portfolio
Assuming an initial investment of $1,000, 
we construct an equal-weighted portfolio 
where each stock receives the same weight ($w_i = 1/N$).
```{r}
initial_investment <- 1000
n <- length(tickers) # Number of stocks in the portfolio
weights <- rep(1 / n, n) # Each stock receives the same weight
```

3. Portfolio Value Calculation (Buy-and-Hold)
We calculate the portfolio value at the end of each month. 
Under a buy-and-hold strategy, the portfolio is not rebalanced, and 
the initial weights remain fixed. We compute the returns using matrix multiplication.
```{r}
# Determine the exact number of shares bought on day 1
first_prices <- as.numeric(prices[1, ])
shares_bought <- (initial_investment * weights) / first_prices

# Calculate the portfolio value
# at the end of each month using matrix multiplication
portfolio_value <- coredata(prices) %*% shares_bought
portfolio_xts <- xts(portfolio_value, order.by = index(prices))
colnames(portfolio_xts) <- "Portfolio"

# Calculate the value of individual stock holdings for comparison plotting
indiv_values <- sweep(coredata(prices), 2, shares_bought, "*")
indiv_xts <- xts(indiv_values, order.by = index(prices))
```

4. Plot Portfolio Performance
Below is the plot of the cumulative return of 
the equal-weighted portfolio over time, alongside the individual stocks.
```{r}
# 1. Ensure the required packages are actively loaded
library(tidyr)
library(ggplot2)
# Combine portfolio and individual stock values
plot_data <- merge(portfolio_xts, indiv_xts)
plot_df <- data.frame(Date = index(plot_data), coredata(plot_data))

# Reshape data to long format for ggplot
plot_df_long <- pivot_longer(
                             plot_df,
                             cols = -Date,
                             names_to = "Asset",
                             values_to = "Value")

# Plot cumulative return over time
ggplot(plot_df_long,
       aes(x = Date,
           y = Value,
           color = Asset,
           linewidth = Asset == "Portfolio")) +
  geom_line() +
  scale_linewidth_manual(values = c("TRUE" = 1.5, "FALSE" = 0.5)) +
  labs(title = "Buy-and-Hold Portfolio Value vs. Individual Stocks",
       subtitle = paste("Initial Investment: $", initial_investment),
       y = "Cumulative Value ($)",
       x = "Date") +
  theme_minimal()
```


## Part 2: Portfolio Optimization
1. Efficient Frontier
We select two stocks from different sectors 
(WMT and NVDA) and use their price data to plot the efficient frontier.
```{r}
# Select two stocks from different sectors and define a custom sample period
tickers_opt <- c("WMT", "NVDA")
getSymbols(tickers_opt,
           src = "yahoo",
           from = "2020-01-01",
           to = "2025-01-01", periodicity = "monthly")

# Merge adjusted prices
prices_opt <- do.call(merge, lapply(tickers_opt, function(x) Ad(get(x))))
colnames(prices_opt) <- tickers_opt
prices_opt <- na.omit(prices_opt)

# Calculate monthly discrete returns
rets_opt <- ROC(prices_opt, type = "discrete")[-1]

# Annualize expected returns and the covariance matrix
mean_rets <- colMeans(rets_opt) * 12
cov_matrix <- cov(rets_opt) * 12

# Generate a sequence of weight combinations for the two stocks (from 0 to 1)
w1 <- seq(0, 1, by = 0.01)
w2 <- 1 - w1

# Calculate expected portfolio
# return and standard deviation for each weight combination
port_rets <- w1 * mean_rets[1] + w2 * mean_rets[2]
port_sds <- sqrt(
                 w1^2 * cov_matrix[1, 1]
                 + w2^2 * cov_matrix[2, 2] + 2 * w1 * w2 * cov_matrix[1, 2])

# Store results in a data frame
frontier_df <- data.frame(Return = port_rets, SD = port_sds, W1 = w1, W2 = w2)
```

2. Minimum-Variance and Optimal Risky Portfolio
Assuming a risk-free rate of 3%, we identify the minimum-variance 
portfolio and the optimal risky portfolio (the tangency portfolio) 
and mark them on the efficient frontier.
```{r}
rf <- 0.03 # Assume a risk-free rate of 3%

# Identify the minimum-variance portfolio
min_var_idx <- which.min(frontier_df$SD)
min_var_port <- frontier_df[min_var_idx, ]

# Determine the optimal risky portfolio
# (tangency portfolio) by maximizing the Sharpe Ratio
sharpe_ratios <- (frontier_df$Return - rf) / frontier_df$SD
opt_risky_idx <- which.max(sharpe_ratios)
opt_risky_port <- frontier_df[opt_risky_idx, ]

# Plot the efficient frontier and mark both portfolios
ggplot(frontier_df, aes(x = SD, y = Return)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_point(
             data = min_var_port,
             aes(x = SD, y = Return),
             color = "red", size = 4) +
  geom_point(data = opt_risky_port,
             aes(x = SD, y = Return),
             color = "green", size = 4) +
  # Draw the Capital Allocation Line (CAL)
  geom_abline(
              intercept = rf,
              slope = max(sharpe_ratios),
              linetype = "dashed", color = "darkgreen") +
  labs(title = "Two-Asset Efficient Frontier (WMT & NVDA)",
       x = "Risk (Annualized Standard Deviation)", 
       y = "Expected Annual Return") +
  annotate("text",
           x = min_var_port$SD,
           y = min_var_port$Return,
           label = "Min-Variance", vjust = -1.5, color="red") +
  annotate("text", x = opt_risky_port$SD,
           y = opt_risky_port$Return,
           label = "Optimal Risky", vjust = -1.5,
           color = "green") +
  theme_minimal()
```