---
title: "Sentiment analysis to 10-K filings"
author: "Huy Le"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction
This document outlines the process of downloading, extracting, and cleaning Item 7 (Management's Discussion & Analysis) from the 10-K filings of five distinct S&P 500 firms for the years 2019 and 2020.

The five selected firms representing different industries are:

Apple Inc. (AAPL) - Technology (CIK: 0000320193)

Johnson & Johnson (JNJ) - Healthcare (CIK: 0000200406)

Exxon Mobil Corp. (XOM) - Energy (CIK: 0000034088)

Walmart Inc. (WMT) - Consumer Staples (CIK: 0000104169)

JPMorgan Chase & Co. (JPM) - Financials (CIK: 0000019617)

1. Load Required Libraries
We will use edgar for fetching the SEC filings, tm for text mining and corpus creation, and stringr/readr for text manipulation.
```{r setup}
# Install packages if not already installed
# install.packages(c("edgar", "tm", "stringr", "readr", "dplyr"))

library(edgar)
library(tm)
library(stringr)
library(readr)
library(dplyr)
```

2. Define Target Firms and Set SEC User Agent
The SEC strictly requires users to declare their User Agent (Name and Email) to prevent server overload. We will also define our target CIK numbers and the years.
```{r}
# SEC EDGAR requires a user agent identifying the scraper
# Update this with your actual name and email
Sys.setenv(EDGAR_USER_AGENT = "Your Name your.email@example.com")

# Define the CIKs for AAPL, JNJ, XOM, WMT, and JPM
target_ciks <- c("320193", "200406", "34088", "104169", "19617")

# Define the target years
target_years <- c(2019, 2020)
```

3. Download and Extract MD&A (Item 7)
The getMgmtDisc function from the edgar package specifically searches for and extracts "Item 7. Management’s Discussion and Analysis of Financial Condition and Results of Operations" from 10-K filings.

Note: This function creates a directory called "MD&A section text" in your working directory and saves the extracted texts as .txt files.
```{r}
# Execute the extraction (Set eval=FALSE in chunk options if already downloaded)
# This will result in 10 text files (5 firms x 2 years)
getMgmtDisc(cik.no = target_ciks, filing.year = target_years)
```

4. Load Extracted Texts into R
After the edgar package downloads and parses the files, we need to read these text files back into R to create our corpus.
```{r}
# Define the directory where the edgar package saves the MD&A files
mda_dir <- "MD&A section text"

# List all text files in the directory
# Make sure the directory exists and contains our 10 target files
if(dir.exists(mda_dir)) {
  file_paths <- list.files(mda_dir, pattern = "\\.txt$", full.names = TRUE)
  
  # Read the content of each file into a character vector
  mda_texts <- sapply(file_paths, function(x) {
    # Read the file and collapse into a single string
    text <- read_file(x)
    return(text)
  }, USE.NAMES = FALSE)
  
  # Create a dataframe to keep track of metadata (using file names)
  file_names <- basename(file_paths)
  docs_df <- data.frame(
    doc_id = file_names,
    text = mda_texts,
    stringsAsFactors = FALSE
  )
  
  print(paste("Successfully loaded", nrow(docs_df), "documents."))
} else {
  warning("The directory 'MD&A section text' does not exist. Please ensure the download step ran successfully.")
}
```

5. Create and Clean the Corpus
Now we convert the text data into a tm package Volatile Corpus (VCorpus) and apply standard NLP cleaning procedures:

Convert everything to lowercase.

Remove punctuation.

Remove numbers.

Remove standard English stopwords (e.g., "the", "and", "is").

Strip extra whitespace.

```{r}
# Ensure the dataframe exists before proceeding
if(exists("docs_df")) {
  
  # Convert the dataframe to a tm Corpus
  mda_corpus <- VCorpus(DataframeSource(docs_df))
  
  # Define a custom function to remove special characters if needed
  removeSpecialChars <- content_transformer(function(x) str_replace_all(x, "[^[:alnum:][:space:]]", " "))
  
  # Apply cleaning transformations
  mda_corpus_clean <- mda_corpus %>%
    tm_map(content_transformer(tolower)) %>%      # Convert to lower case
    tm_map(removeSpecialChars) %>%                # Remove special characters
    tm_map(removePunctuation) %>%                 # Remove punctuation
    tm_map(removeNumbers) %>%                     # Remove numbers
    tm_map(removeWords, stopwords("en")) %>%      # Remove standard English stopwords
    tm_map(stripWhitespace)                       # Remove extra white spaces
  
  # Inspect the first document's metadata and a snippet of its clean text
  print(meta(mda_corpus_clean[[1]]))
  print(substr(content(mda_corpus_clean[[1]]), 1, 500)) # Print first 500 characters
  
}
```

Conclusion
The 10 documents have been successfully downloaded, extracted, and processed. The variable mda_corpus_clean now holds the fully cleaned text corpus, which is ready for downstream text analytics tasks such as sentiment analysis, topic modeling (e.g., LDA), or calculating term frequency-inverse document frequency (TF-IDF).

