---
title: "Yahoo Finance Intro"
author: "JQ"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

## Overview
This session covers:

- Import stock price data from Yahoo Finance.
- Inspect and understand xts objects.
- Perform basic data manipulation.
- Normalize prices.
- Plot capital appreciation of multiple securities.


---

## Import stock price data from Yahoo Finance

```{r setup, message=FALSE, warning=FALSE}
library(quantmod)  # package for financial data and time-series tools

# Download daily data for NVDA from Yahoo Finance
getSymbols("NVDA", 
           from = "2019-12-31", 
           to   = "2024-12-31", 
           auto.assign = TRUE)

head(NVDA)         # preview first few rows
class(NVDA)        # check object class
```

### What is xts?

- An extensible time-series object built on **zoo**.
- Rows are indexed by time (Date or POSIXct).
- Very efficient for financial data operations (merging, subsetting by date, rolling windows).

### Common Pitfalls

- Avoid using `$` with xts objects; use `[, "ColumnName"]` instead.
- Always check the index class before time-based subsetting.


---

## Checking the Data Structure

```{r}
head(NVDA, 3)      # first 3 observations
tail(NVDA, 3)      # last 3 observations
dim(NVDA)          # number of rows and columns
summary(NVDA)      # summary statistics
```

```{r}
# Plot closing price
plot(NVDA[, "NVDA.Close"],
     main = "NVDA Closing Price",
     col = "darkblue",
     lwd = 2)
```

---

## Basic Data Manipulation

```{r}
NVDA[1, ]              # first observation
NVDA[1:5, ]            # first 5 observations
nrow(NVDA)             # total number of rows

# Select specific rows
NVDA[c(1, 251), ]

# Extract only Close price column
close_price <- NVDA[, "NVDA.Close"]
head(close_price)

# Date-based subsetting (January 2024 only)
NVDA_2024Jan <- NVDA["2024-01"]
head(NVDA_2024Jan)

# Logical filtering example
high_volume_days <- subset(NVDA,
                           NVDA.Volume > 500000000 &
                           index(NVDA) >= as.Date("2024-05-01"))
head(high_volume_days)
```

---

## Convert to Weekly Prices

```{r}
# Convert daily data to weekly frequency
NVDA_weekly <- to.weekly(NVDA)

head(NVDA_weekly)   # weekly OHLC data
```

---

## Compare Capital Gains Across Multiple Securities

### Download Multiple Stocks

```{r}
symbols <- c("NVDA", "TSLA", "GOOG", "SPY")

getSymbols(symbols,
           from = "2019-12-31",
           to   = "2024-12-31")
```

---

## Combine and Normalize Prices

```{r}
# Combine adjusted closing prices
close.prc <- cbind(NVDA[, "NVDA.Close"],
                   TSLA[, "TSLA.Close"],
                   GOOG[, "GOOG.Close"],
                   SPY[,  "SPY.Close"])

colnames(close.prc) <- symbols

head(close.prc)

# Normalize prices to start at 1 (Value of $1 investment)
norm.prc <- close.prc
norm.prc[, "NVDA"] <- close.prc[, "NVDA"] / as.numeric(close.prc[1, "NVDA"])
norm.prc[, "TSLA"] <- close.prc[, "TSLA"] / as.numeric(close.prc[1, "TSLA"])
norm.prc[, "GOOG"] <- close.prc[, "GOOG"] / as.numeric(close.prc[1, "GOOG"])
norm.prc[, "SPY"] <- close.prc[, "SPY"] / as.numeric(close.prc[1, "SPY"])
head(norm.prc)
```

---

## Plot Capital Appreciation

```{r}
y.range <- range(norm.prc)     # determine y-axis range
dt <- index(norm.prc)          # extract date index

plot(dt, norm.prc[, "NVDA"],
     type = "l",
     lwd = 2,
     col = "black",
     ylim = y.range,
     xlab = "Date",
     ylab = "Value of $1 Investment",
     main = "Value of $1 Investment: NVDA, TSLA, GOOG, SPY")

lines(dt, norm.prc[, "TSLA"], col = "blue")
lines(dt, norm.prc[, "GOOG"], col = "red")
lines(dt, norm.prc[, "SPY"],  col = "darkgreen")

abline(h = 1, lty = 2)  # reference line at initial value

legend("topleft",
       legend = symbols,
       col = c("black", "blue", "red", "darkgreen"),
       lwd = 2)
```

---

## Key Takeaways

- `quantmod` makes financial data import easy.
- `xts` objects allow powerful time-based subsetting.
- Normalizing prices allows meaningful performance comparison.
- Always use adjusted or closing prices consistently.
- Visualizing cumulative returns helps interpret capital appreciation.

---
